---
title: "NHANES Public Health Data Analysis"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show R code in the output
  message = FALSE,     # Suppress package messages
  warning = FALSE,     # Suppress warnings
  comment = "#>"       # Prefix for output lines
)
```

NHANES is a program run by the CDC to assess the health and nutritional
status of adults and children in the US. It combines survey questions
and physical examinations, including medical and physiological
measurements and laboratory tests, and examines a representative sample
of about 5,000 people each year. The data is used to determine the
prevalence of diseases and risk factors, establish national standards,
and support epidemiology studies and health sciences research. This
information helps to develop public health policy, design health
programs and services, and expand the nation's health knowledge.

You can find those data using these links:

**Link for the Data**: <https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv>

**Dictionary Data**: <https://github.com/HackBio-Internship/public_datasets/blob/main/R/nhanes_dd.csv>

**Tasks**:

• Process all NA (either by deleting or by converting to zero)

• Visualize the distribution of BMI, Weight, Weight in pounds (weight
\*2.2) and Age with an histogram.

• Determine the mean 60-second pulse rate for all participants in the
data?

• What’s the range of values for diastolic blood pressure in all
participants? (Hint: see help for min(), max()).

• What’s the variance and standard deviation for income among all
participants?

• Visualize the relationship between weight and height. Color the points
by:

o gender

o diabetes

o smoking status

• Conduct t-test between the following variables and make conclusions on
the relationship between them based on P-Value

o Age and Gender

o BMI and Diabetes

o Alcohol Year and Relationship Status

# Load necessary packages

```{r Packages}

```

