---
title: "NHANES Public Health Data Analysis"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show R code in the output
  message = FALSE,     # Suppress package messages
  warning = FALSE,     # Suppress warnings
  comment = "#>"       # Prefix for output lines
)
```

NHANES is a program run by the CDC to assess the health and nutritional
status of adults and children in the US. It combines survey questions
and physical examinations, including medical and physiological
measurements and laboratory tests, and examines a representative sample
of about 5,000 people each year. The data is used to determine the
prevalence of diseases and risk factors, establish national standards,
and support epidemiology studies and health sciences research. This
information helps to develop public health policy, design health
programs and services, and expand the nation's health knowledge.

You can find those data using these links:

**Link for the Data**: <https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv>

**Dictionary Data**: <https://github.com/HackBio-Internship/public_datasets/blob/main/R/nhanes_dd.csv>

**Tasks**:

• Process all NA (either by deleting or by converting to zero)

• Visualize the distribution of BMI, Weight, Weight in pounds (weight
\*2.2) and Age with an histogram.

• Determine the mean 60-second pulse rate for all participants in the
data?

• What’s the range of values for diastolic blood pressure in all
participants? (Hint: see help for min(), max()).

• What’s the variance and standard deviation for income among all
participants?

• Visualize the relationship between weight and height. Color the points
by:

o gender

o diabetes

o smoking status

• Conduct t-test between the following variables and make conclusions on
the relationship between them based on P-Value

o Age and Gender

o BMI and Diabetes

o Alcohol Year and Relationship Status

# 1. Load necessary packages

```{r}
# Install if not already installed
if (!require("pacman")) install.packages("pacman")

# Load all packages
pacman::p_load(
  tidyverse,   # data manipulation + ggplot2
  patchwork,   # combining plots
  scales       # axis formatting
)
```

```{r Packages}
library(tidyverse)
library(patchwork)
library(scales)
```

# 2. Load the Data and Present the Data Dictionnary

```{r}
# Source: CDC NHANES via HackBio public datasets repository
# Each row = one survey participant (~5,000 sampled per year)

# ── Load the data ──────────────
nhanes_raw <- read.csv(
  "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv",
  stringsAsFactors = FALSE   # keep character columns as strings, not factors
)

# ── Create the data dictionary as a data frame ──────────────
nhanes_dict <- data.frame(
  Column = c(
    "id", "Gender", "Age", "Race", "Education",
    "MaritalStatus", "RelationshipStatus", "Insured", "Income",
    "Poverty", "HomeRooms", "HomeOwn", "Work", "Weight", "Height",
    "BMI", "Pulse", "BPSys", "BPDia", "Testosterone",
    "HDLChol", "TotChol", "Diabetes", "DiabetesAge",
    "nPregnancies", "nBabies", "SleepHrsNight", "PhysActive",
    "PhysActiveDays", "AlcoholDay", "AlcoholYear", "SmokingStatus"
  ),
  Type = c(
    "integer",   "character", "integer",   "character", "character",
    "character", "character", "character", "numeric",
    "numeric",   "integer",   "character", "character", "numeric",  "numeric",
    "numeric",   "integer",   "integer",   "integer",   "numeric",
    "numeric",   "numeric",   "character", "integer",
    "integer",   "integer",   "integer",   "character",
    "integer",   "numeric",   "integer",   "character"
  ),
  Label = c(
    "Unique participant ID",
    "Biological sex of participant (male / female)",
    "Age in years at screening",
    "Race / ethnicity of participant",
    "Highest education level attained",
    "Current marital status",
    "Relationship status (Committed / Single)",
    "Whether participant has health insurance (Yes / No)",
    "Annual gross household income in USD",
    "Ratio of household income to poverty threshold",
    "Number of rooms in the participant's home",
    "Home ownership status (Own / Rent / Other)",
    "Current employment status",
    "Body weight in kilograms",
    "Standing height in centimetres",
    "Body Mass Index — weight(kg) / height(m)²",
    "60-second resting pulse rate (beats per minute)",
    "Systolic blood pressure (mmHg)",
    "Diastolic blood pressure (mmHg)",
    "Serum testosterone level (ng/dL)",
    "HDL (good) cholesterol level (mmol/L)",
    "Total cholesterol level (mmol/L)",
    "Whether participant was diagnosed with diabetes (Yes / No)",
    "Age at diabetes diagnosis (years)",
    "Total number of pregnancies",
    "Number of live births",
    "Average hours of sleep per night",
    "Whether participant is physically active (Yes / No)",
    "Number of days per week with physical activity",
    "Average number of alcoholic drinks on drinking days",
    "Number of days per year alcohol was consumed",
    "Smoking history (Never / Former / Current)"
  ),
  stringsAsFactors = FALSE
)

# ── View the full dictionary ─────────────────────────────────
print(nhanes_dict)

# ── Or view it as a clean table in RStudio Viewer ───────────
View(nhanes_dict)

# ── Quick lookup function — get label for any column name ───
get_label <- function(col_name) {
  match <- nhanes_dict$Label[nhanes_dict$Column == col_name]
  if (length(match) == 0) return("Column not found")
  return(match)
}

# Example usage:
get_label("BMI")          # "Body Mass Index — weight(kg) / height(m)²"
get_label("AlcoholYear")  # "Number of days per year alcohol was consumed"
get_label("BPDia")        # "Diastolic blood pressure (mmHg)"
get_label("Work")
```
# 3. Quick Overview of the data

```{r}
# Keep a working copy , never modify the raw object directly
nhanes <- nhanes_raw

# ── Structure of the dataset ─────────────────────────────────
# Shows column names, data types, and first values
str(nhanes)

# ── Name of all Variables ─────────────────────────────────
names(nhanes)

# ── Unique values in key categorical columns ─────────────────
# Useful to spot typos, unexpected categories, or hidden NAs
unique(nhanes$Gender)
unique(nhanes$Race)
unique(nhanes$Education)
unique(nhanes$MaritalStatus)
unique(nhanes$RelationshipStatus)
unique(nhanes$Diabetes)
unique(nhanes$SmokingStatus)
unique(nhanes$Work)
unique(nhanes$HomeOwn)

# ── First and last rows ──────────────────────────────────────
head(nhanes)          # first 6 rows
tail(nhanes, n = 10)  # last 10 rows

# ── Dimensions ───────────────────────────────────────────────
# How many participants (rows) and variables (columns)?
dim(nhanes)

# ── Detailed column-by-column overview ───────────────────────
# Shows data type and first values for every column
glimpse(nhanes)

# ── Statistical summary ──────────────────────────────────────
# Numeric columns → min, Q1, median, mean, Q3, max, NA count
# Character columns → length and class
summary(nhanes)

# ── Missing values per column ────────────────────────────────
# Shows exact NA count and percentage for each column

colSums(is.na(nhanes_raw))

na_count <- colSums(is.na(nhanes))
na_pct   <- round(na_count / nrow(nhanes) * 100, 2)

na_summary <- data.frame(
  Column       = names(na_count),
  Missing      = na_count,
  Percentage   = paste0(na_pct, "%")
)

na_summary

# Show only columns that actually have missing values
na_summary[na_summary$Missing > 0, ]
```

Based on the `na_summary` output from our code, here is what we can retain:


**Columns with NO missing values** — complete data for all participants:
- `id`, `Gender`, `Age`, `Race`
- These columns are **reliable for all analyses**

---

**Columns with MODERATE missingness** — present but incomplete:
- `HomeRoom`, `HomeOwn`: Missing because some `Participants refused to answer`
- `Education`, `MaritalStatus`, `RelationshipStatus`, `Work`,  → missing for **younger participants** (children don't have jobs or marital status)
- `BMI`, `Weight`, `Height`, `Pulse`, `BPSys`, `BPDia` → missing because some **physical exams were not completed**
- `Income`, `Poverty` → missing because some participants **refused to answer**
- `HDLChol`, `TotChol`, `Testosterone` → missing because **lab tests were not done** for everyone

---

**Columns with HIGH missingness** — structurally incomplete:
- `DiabetesAge` → only filled when `Diabetes == "Yes"`, so **~90% are NA** naturally
- `nPregnancies`, `nBabies` → only applicable to **female participants**
- `AlcoholDay`, `AlcoholYear` → only filled for participants who **reported drinking**
- `PhysActiveDays` → only filled for participants who said **Yes to PhysActive**

---

**Key conclusions to retain:**

> **1.** Most NAs are **not random errors** — they are structurally expected because certain questions only apply to subgroups (females, diabetics, drinkers, etc.)

> **2.** Replacing ALL NAs with `0` makes sense for columns like `AlcoholYear` (non-drinkers genuinely consumed 0 days) and `nBabies` (males had 0 births), but would be **misleading** for columns like `Income` or `BPDia`

> **3.** For statistical tests like our **t-tests**, R's built-in `na.rm = TRUE` argument handles remaining NAs gracefully — so we don't need to delete entire rows

> **4.** Deleting all rows with ANY NA would **destroy most of the dataset**, since almost every participant is missing at least one value

# 4 Treatment of Missing data

Let us solve each missing value per columns 

##1. HomeRooms and HomeOwn

28 Participants were enable to give an answer. We have 28 missing values fpr both `HoneRooms` and `HomeOwn`. We will replace missing values for these two columns by `No answer`

```{r}
unique(nhanes$HomeRooms)
# show lines where we have only misisng values related to these two columns
missing_count_1 <- data.frame(
  Variable = c("HomeRooms", "HomeOwn"),
  Missing  = c(sum(is.na(nhanes$HomeRooms)),
               sum(is.na(nhanes$HomeOwn)))
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
  )

missing_count_1

nhanes %>%
  filter(is.na(HomeRooms)
         
# For realability, we created a function
         
filter_na <- function(data, column) {
  data %>%
    filter(is.na({{ column }}))
}

# Combinaison of Both 
filter_na(nhanes, HomeRooms)

nhanes %>%
  filter(is.na(HomeRooms) | is.na(HomeOwn))

         
# Transform the variable type for this two data in factor 

nhanes<- nhanes %>%
  mutate(
    HomeRooms= factor(HomeRooms),
    HomeOwn = factor(HomeOwn)
  )
```

```{r}

# Remplacement of missing values by `No asnwer`

nhanes <- nhanes %>%
  mutate(across(c(HomeRooms, HomeOwn),
                ~forcats::fct_explicit_na(., na_level = "Unknown")))

#or

nhanes <- nhanes %>%
  mutate(across(c(HomeRooms, HomeOwn),
                ~replace_na(as.character(.), "Unknown")))


nhanes <- nhanes %>%
  mutate(
    HomeRooms = replace_na(HomeRooms, "Unknown"),
    HomeOwn = replace_na(HomeOwn, "Unknown")
  ) # don't work because HomeRooms is a factor variable. I am trying to include new factor which doesn't exist before.

unique(nhanes$HomeOwn)

summary(nhanes)

str(nhanes$HomeRooms)


nhanes <- nhanes %>%
  mutate(across(c(HomeRooms, HomeOwn), ~replace_na(., "Unknown")))

```

##2.  Education, MaritalStatus, RelationshipStatus and Work

We can see that we have missing values for **younger participants** (children don't have jobs or marital status)

We can manage missing values by including the `no_asnwer` statement

```{r}

# visualisation of the strcuture of each variables
str(nhanes$Education)
str(nhanes$MaritalStatus)
str(nhanes$RelationshipStatus)
str(nhanes$Work)

unique(nhanes$Education)
unique(nhanes$MaritalStatus)
unique(nhanes$RelationshipStatus)
unique(nhanes$Work)

# coount of missing values

missing_count_2 <- data.frame(
  Variable = c("Education", "MaritalStatus", "RelationshipStatus", "work"),
  Missing  = c(sum(is.na(nhanes$Education)),
               sum(is.na(nhanes$MaritalStatus)),
               sum(is.na(nhanes$RelationshipStatus)),
               sum(is.na(nhanes$Work)))
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
  )

missing_count_2

# determination of each rows with missing values for the variables Education, Marrital status, Relationship status and work

## Education
nhanes %>%
  filter(is.na(Education) # Majority of young individuals doesn't give an answer to this question

#or

filter_na(nhanes, Education)

## Marrital Status

nhanes %>%
  filter(is.na(MaritalStatus) # Majority of young individuals doesn't give an answer to this question

#or

filter_na(nhanes, MaritalStatus)

## Relationship Status

nhanes %>%
  filter(is.na(RelationshipStatus) # Majority of young individuals doesn't give an answer to this question

#or

filter_na(nhanes, RelationshipStatus)

##Work

nhanes %>%
  filter(is.na(Work) # Majority of young individuals doesn't give an answer to this question

#or

filter_na(nhanes, Work)

##  Combinaison of both

nhanes %>%
  filter(is.na(Education) | is.na(MaritalStatus) | is.na(RelationshipStatus) | is.na(Work))

```

```{r}
# Remplacement of missing values by `No asnwer`

nhanes <- nhanes %>%
  mutate(across(c(Education, MaritalStatus, RelationshipStatus, Work),
                ~forcats::fct_explicit_na(., na_level = "Unknown")))

# verification

unique(nhanes$Education)
unique(nhanes$MaritalStatus)
unique(nhanes$RelationshipStatus)
unique(nhanes$Work)

```

```{r}
# Transform the variable type for this two data in factor 

nhanes<- nhanes %>%
  mutate(
    Education = factor(Education),
    MaritalStatus = factor(MaritalStatus),
    RelationshipStatus = factor(RelationshipStatus),
    Work = factor(Work)
  )

```


##3. BMI, Weight, Height, Pulse, BPSys, BPDia

Here, we have missing because some physical exams were not completed

```{r}
# visualisation of the strcuture of each variables
str(nhanes$BMI)
str(nhanes$Weight)
str(nhanes$Height)
str(nhanes$Pulse)
str(nhanes$BPSys)
str(nhanes$BPDia)

summary(nhanes$BMI)
summary(nhanes$Weight)
summary(nhanes$Height)
summary(nhanes$Pulse)
Summary(nhanes$BPSys)
summary(nhanes$BPDia)

# coount of missing values

missing_count_3 <- data.frame(
  Variable = c("BMI", "Weight", "Height", "Pulse", "BPSys", "BPDia"),
  Missing  = c(sum(is.na(nhanes$BMI)),
               sum(is.na(nhanes$Weight)),
               sum(is.na(nhanes$Height)),
               sum(is.na(nhanes$Pulse)),
               sum(is.na(nhanes$BPSys)),
               sum(is.na(nhanes$BPDia)))
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
  )

missing_count_3

# determination of each rows with missing values for the variables BMI, Weight, Height, Pulse, BPSys, BPDia

## BMI
nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(BMI)) # Majority of young individuals doesn't have this measures
#or

filter_na(nhanes, BMI)

## Weight

nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(Weight))

#or

filter_na(nhanes, Weight)

## Height

nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(Height))

#or

filter_na(nhanes, Height)

##Pulse

nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(Pulse)) # Majority of young individuals doesn't give an answer to this question

#or

filter_na(nhanes, Pulse)

## BPSys

nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(BPSys))

#or

filter_na(nhanes, BPSys)

## BPDia

nhanes %>%
  select(id, Gender, Age, BMI, Weight, Height, Pulse, BPSys, BPDia) %>% 
  filter(is.na(BPDia))

# OR

filter_na(nhanes, BPDia)

##  Combinaison of both

nhanes %>%
  filter(is.na(BMI) | is.na(Weight) | is.na(Height) | is.na(Pulse) | is.na(BPDia) | is.na(BPSys)) # or

nhanes %>%
  filter(is.na(BMI) & is.na(Weight) & is.na(Height) & is.na(Pulse) & is.na(BPDia) & is.na(BPSys))

nhanes %>%
  filter(is.na(Pulse) & is.na(BPDia) & is.na(BPSys))
```


We can retain that

> **1.** Most NAs for these variables are **not random errors**: children between 0 to 1 years old don't have reuslts or `Pulse`, `BPDia` and `BPSys`
To resolve missinsg values, we will replace NAs by 0

> **2.** For `BMI`, `Height` and `Weight`, we notice the same phenomena. Majprity of participants without `Height` and `BMI` values are young children. We will replace missing values by zero. 

```{r}

# Remplacement of missing values by `0`

nhanes <- nhanes %>% 
  mutate(across(
    c(BMI, Height, Weight, Pulse, BPDia, BPSys),
    ~ifelse(is.na(.), 0, .) # replace all misisng values by zero
  )) %>% 
  ungroup()

# it was even better to alloow misising values and don't considered the lines with NAs during statistical analysis.

str(nhanes$BMI)
str(nhanes$Weight)
str(nhanes$Height)
str(nhanes$Pulse)
str(nhanes$BPSys)
str(nhanes$BPDia)

summary(nhanes$BMI)
summary(nhanes$Weight)
summary(nhanes$Height)
summary(nhanes$Pulse)
Summary(nhanes$BPSys)
summary(nhanes$BPDia)
```
##4.Income and Poverty

Here, we can conclude that we have missing values because participants refused to give an answer. Poverty depend on the income.If someone refused to give his Income, we can conclude on his social life. Also, young children and some adults don't have an Income since they are not working. The income value depend also on the age of the participants. Yound and adults doesn't have an income.  

```{r}
# visualisation of the structure of each variables

summary(nhanes$Income)
summary(nhanes$Poverty)

class(nhanes$Income)

# count of missing values

missing_count_4 <- data.frame(
  Variable = c("Income", "Poverty"),
  Missing  = c(sum(is.na(nhanes$Income)),
               sum(is.na(nhanes$Poverty)))
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
)

missing_count_4

# determination of each rows with missing values for the variables Income and Poverty 

## Income

nhanes %>%
  select(id, Gender, Age, Income, Poverty) %>% 
  filter(is.na(Income)) # Majority of young individuals and old peoples doesn't have this measures
#or

filter_na(nhanes, Income)

## Poverty


nhanes %>%
  select(id, Gender, Age, Income, Poverty) %>% 
  filter(is.na(Poverty))

#or

filter_na(nhanes, Poverty)

##  Combinaison of both

nhanes %>%
  filter(is.na(Income) | is.na(Poverty)) # or

nhanes %>%
  filter(is.na(Income) & is.na(Poverty)) # both
```

> We will replace missing values here by zero

```{r}
# Remplacement of missing values by `0`

nhanes <- nhanes %>% 
  mutate(across(
    c(Income, Poverty),
    ~ifelse(is.na(.), 0, .) # replace all misisng values by zero. It can be misleasing to replace income andpoverty by zero, since we don't know if the persone has a job or not. But based on the agee of the participant, we conlued to replace Nas by zero.
  )) %>% 
  ungroup()

str(nhanes$Income)
str(nhanes$Poverty)

summary(nhanes$Income)
summary(nhanes$Poverty)
```

##5. HDLChol, TotChol, Testosterone

Here, we have missing values because **lab tests were not done** for everyone. Also, `Testoterone` was measured only for men, not for women, but some Men don't have this value also.  

we have the same number of missing values for `HDL` and `TotChol` (Na = 775), prove that these parameters were not measured at the laboratory due to certain raisons (low Volume of sample available, incapacity to do the test). Majority of participants without `HDL` and `Totchol` levels, don't have also results for Testosterone. why???

```{r}
# visualisation of the structure of each variables

summary(nhanes$HDLChol)
summary(nhanes$TotChol)
summary(nhanes$Testosterone)


# count of missing values

missing_count_5 <- data.frame(
  Variable = c("HDLChol", "TotChol", "Testosterone"),
  Missing  = c(sum(is.na(nhanes$HDLChol)),
               sum(is.na(nhanes$TotChol)),
               sum(is.na(nhanes$Testosterone)))
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
)

missing_count_5

# determination of each rows with missing values for the variables Income and Poverty 

## HDLChol

nhanes %>%
  select(id, Gender, Age, HDLChol, TotChol, Testosterone) %>% 
  filter(is.na(HDLChol)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, HDLChol)

## TotChol

nhanes %>%
  select(id, Gender, Age, HDLChol, TotChol, Testosterone) %>% 
  filter(is.na(TotChol)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, TotChol)


##  Testosterone

nhanes %>%
  select(id, Gender, Age, HDLChol, TotChol, Testosterone) %>% 
  filter(is.na(Testosterone)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, Testosterone)

##  Combination of both

nhanes %>%
  filter(is.na(HDLChol) | is.na(TotChol) | is.na(Testosterone)) # or

nhanes %>%
  filter(is.na(HDLChol) & is.na(TotChol) & is.na(Testosterone)) # both
```

```{r}
# Remplacement of missing values by `0`

nhanes <- nhanes %>% 
  mutate(across(
    c(HDLChol, TotChol, Testosterone),
    ~ifelse(is.na(.),  median(., na.rm = TRUE), .)  # Here, we replace NAs by the median. buut it is better to left NAs value and ignore them duirng statsitical analysis. 
  )) %>% 
  ungroup()

str(nhanes$HDLChol)
str(nhanes$TotChol)
str(nhanes$Testosterone)

summary(nhanes$HDLChol)
summary(nhanes$TotChol)
summary(nhanes$Testosterone)
```
##6. Diabetes and DiabetesAge

We can see that `DiabetesAge` was given when `Diabetes == "Yes"`. We will replace missing values by a `zero`
in  this dataser, we have mor eyounger individuals compared to olders

```{r}
# visualisation of the structure of each variables

summary(nhanes$Diabetes)
str(nhanes$Diabetes)
unique(nhanes$Diabetes)
summary(nhanes$DiabetesAge)

nhanes %>% 
  count(Age, name = "nombre_participants") %>% 
  arrange(Age) 

nhanes %>% 
  count(Diabetes, name = "nombre_participants")

nhanes %>% 
  count(DiabetesAge, name = "nombre_participants")

# count of missing values

missing_count_6 <- data.frame(
  Variable = c("Diabetes", "DiabetesAge"),
  Missing  = c(sum(is.na(nhanes$Diabetes)),
               sum(is.na(nhanes$DiabetesAge)))
              
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
)

missing_count_6

# determination of each rows with missing values for the variables Income and Poverty 

## Diabetes

nhanes %>%
  select(id, Gender, Age, Diabetes, DiabetesAge) %>% 
  filter(is.na(Diabetes)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, Diabetes)

## Diabetes

nhanes %>%
  select(id, Gender, Age, Diabetes, DiabetesAge) %>% 
  filter(is.na(DiabetesAge)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, DiabetesAge)

##  Combination of both

nhanes %>%
  filter(is.na(Diabetes) | is.na(DiabetesAge)) # or

nhanes %>%
  filter(is.na(Diabetes) & is.na(DiabetesAge)) # both
```

```{r}
# Replacement of missing values for Diabetes by no_answer

nhanes <- nhanes %>%
  mutate(across(c(Diabetes),
                ~forcats::fct_explicit_na(., na_level = "Unknown")))

# Replacement of missing Values for DiabetesAge by zero

nhanes <- nhanes %>% 
  mutate(across(
    c(DiabetesAge),
    ~ifelse(is.na(.),  0, .)  # Here, we replace NAs by 0.  but it is better to left NAs value and ignore them during statistical analysis. 
  )) %>% 
  ungroup()

str(nhanes$DiabetesAge)
str(nhanes$Diabetes)


summary(nhanes$Diabetes)
summary(nhanes$DiabetesAge)
```

##7 nPregnancies and nBabies

These are only applicable to **female participants**.  We will replace missing values by a zero answer

```{r}
# visualisation of the structure of each variables

summary(nhanes$nPregnancies)
summary(nhanes$nBabies)

str(nhanes$nPregnancies)
str(nhanes$nBabies)

nhanes %>% 
  count(nPregnancies, name = "nombre_participants") %>% 
  arrange(nPregnancies) 

nhanes %>% 
  count(nBabies, name = "nombre_participants")

nhanes %>% 
  count(DiabetesAge, name = "nombre_participants")

# count of missing values

missing_count_7 <- data.frame(
  Variable = c("nPregnancies", "nBabibies"),
  Missing  = c(sum(is.na(nhanes$nPregnancies)),
               sum(is.na(nhanes$nBabies)))
              
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
)

missing_count_7

# determination of each rows with missing values for the variables Income and Poverty 

## Diabetes

nhanes %>%
  select(id, Gender, Age, nPregnancies, nBabies) %>% 
  filter(is.na(nPregnancies)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, nPregnancies)

## Diabetes

nhanes %>%
  select(id, Gender, Age, nPregnancies, nBabies) %>% 
  filter(is.na(nBabies)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, nBabies)

##  Combination of both

nhanes %>%
  filter(is.na(nPregnancies) | is.na(nBabies)) # or

nhanes %>%
  select(id, Gender, Age, nPregnancies, nBabies) %>% 
  filter(is.na(nPregnancies) | is.na(nBabies))# or

nhanes %>%
  select(id, Gender, Age, nPregnancies, nBabies) %>%
  filter(is.na(nPregnancies) & is.na(nBabies)) # both
```

```{r}
# Replacement of missing Values by zero

nhanes <- nhanes %>% 
  mutate(across(
    c(nPregnancies, nBabies),
    ~ifelse(is.na(.),  0, .)  # Here, we replace NAs by 0.  b
  )) %>% 
  ungroup()

summary(nhanes$nPregnancies)
summary(nhanes$nBabies)
```
##8. AlcoholDay and AlcoholYear

only filled for participants who **reported drinking**. We will replace missing values by a zero

```{r}
# visualisation of the structure of each variables

summary(nhanes$AlcoholDay)
summary(nhanes$AlcoholYear)

nhanes %>% 
  count(AlcoholDay, name = "nombre_participants") %>% 
  arrange(AlcoholDay) 

nhanes %>% 
  count(AlcoholYear, name = "nombre_participants")

# count of missing values

missing_count_8 <- data.frame(
  Variable = c("AlcoholDay", "AlcoholYear"),
  Missing  = c(sum(is.na(nhanes$AlcoholDay)),
               sum(is.na(nhanes$AlcoholYear)))
              
) %>%
  mutate(
    Present    = nrow(nhanes) - Missing,
    Pct_Missing = round(Missing / nrow(nhanes) * 100, 2)
)

missing_count_8

# determination of each rows with missing values for the variables Income and Poverty 

## AlcoholDay

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>% 
  filter(is.na(AlcoholDay)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, AlcoholDay)

## AlcoholYear

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>% 
  filter(is.na(AlcoholYear)) # Majority of young individuals and old peoples doesn't have this measures

#or

filter_na(nhanes, AlcoholYear)

##  Combination of both

nhanes %>%
  filter(is.na(AlcoholDay) | is.na(AlcoholYear)) # or

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>% 
  filter(is.na(AlcoholDay) | is.na(AlcoholYear))# or

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>%
  filter(is.na(AlcoholDay) & is.na(AlcoholYear)) # both
```

```{r}
# Replacement of missing Values by zero

nhanes <- nhanes %>% 
  mutate(across(
    c(AlcoholDay, AlcoholYear),
    ~ifelse(is.na(.),  0, .)  # Here, we replace NAs by 0.  b
  )) %>% 
  ungroup()

summary(nhanes$AlcoholDay)
summary(nhanes$AlcoholYear)
```


##9. PhysActiveDays

only filled for participants who said **Yes to PhysActive**. We will replace missing values by a zero 

```{r}
# visualisation of the structure of each variables

summary(nhanes$PhysActiveDays)

nhanes %>% 
  count(PhysActiveDays, name = "nombre_participants") %>% 
  arrange(PhysActiveDays)


##  Combination of both

nhanes %>%
  filter(is.na(AlcoholDay) | is.na(AlcoholYear)) # or

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>% 
  filter(is.na(AlcoholDay) | is.na(AlcoholYear))# or

nhanes %>%
  select(id, Gender, Age, AlcoholDay, AlcoholYear) %>%
  filter(is.na(AlcoholDay) & is.na(AlcoholYear)) # both

```

```{r}
# Replacement of missing Values by zero

nhanes <- nhanes %>% 
  mutate(across(
    c(PhysActiveDays),
    ~ifelse(is.na(.),  0, .)  # Here, we replace NAs by 0.  b
  )) %>% 
  ungroup()

# Replacement of missing values for Diabetes by no_answer

nhanes <- nhanes %>%
  mutate(across(c(PhysActive),
                ~forcats::fct_explicit_na(., na_level = "Unknown")))

summary(nhanes$PhysActiveDays)
summary(nhanes$PhysActive)
```

```{r}
print(colSums(is.na(nhanes)))

# for sleeping status

nhanes <- nhanes %>%
  mutate(across(c(SmokingStatus),
                ~forcats::fct_explicit_na(., na_level = "Unknown")))

# for sleeping night
nhanes <- nhanes %>% 
  mutate(across(
    c(SleepHrsNight),
    ~ifelse(is.na(.),  0, .)  # Here, we replace NAs by 0.  b
  )) %>% 
  ungroup()
```


#3. creation of a new column 

```{r}
# Add a derived column: Weight in pounds
nhanes <- nhanes %>%
  mutate(Weight_lbs = Weight * 2.2)

dim(nhanes)

cat("\n=== CLEANED DATA — remaining NAs ===\n")
print(colSums(is.na(nhanes)))
```

```{r}
## save this dataset from R

write.csv(nhanes, "nhanes_cleaned.csv", row.names = FALSE)
```

# 4. Let us now visualise the distribution of BMI, Weight, Weight in pounds (weight *2.2) and Age with an histogram.

## 4.1. Distribution of BMI

```{r}
theme_nhanes <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank()
  )

h1 <- ggplot(nhanes, aes(x = BMI)) +
  geom_histogram(bins = 40, 
                 fill = "#2196F3", 
                 color = "white", 
                 alpha = 0.85) +
  labs(title = "Distribution of BMI",
       subtitle = "Body Mass Index (kg/m²)",
       x = "BMI", y = "Count") +
  theme_nhanes

h1
```

## 4.2 Distribution of Weight

```{r}
h2 <- ggplot(nhanes, aes(x = Weight)) +
  geom_histogram(bins = 40, 
                 fill = "#4CAF50", 
                 color = "white", 
                 alpha = 0.85) +
  labs(title = "Distribution of Weight",
       subtitle = "Weight in kilograms",
       x = "Weight (kg)", y = "Count") +
  theme_nhanes

h2
```

## 4.3 distribution of Weight in pounds (weight *2.2)

```{r}
h3 <- ggplot(nhanes, aes(x = Weight_lbs)) +
  geom_histogram(bins = 40,
                 fill = "#FF9800", 
                 color = "white",
                 alpha = 0.85) +
  labs(title = "Distribution of Weight (lbs)",
       subtitle = "Weight in pounds (kg × 2.2)",
       x = "Weight (lbs)", y = "Count") +
  theme_nhanes

h3
```


## 4.4 Distribution of Age

```{r}
h4 <- ggplot(nhanes, aes(x = Age)) +
  geom_histogram(bins = 40, fill = "#9C27B0", color = "white", alpha = 0.85) +
  labs(title = "Distribution of Age",
       subtitle = "Age in years",
       x = "Age (years)", y = "Count") +
  theme_nhanes

h4
```

```{r}
hist_combined <- (h1 | h2) / (h3 | h4) +
  plot_annotation(
    title   = "NHANES — Distribution of Key Variables",
    caption = "Source: CDC NHANES Dataset",
    theme   = theme(plot.title = element_text(face = "bold", size = 16))
  )

print(hist_combined)
ggsave("nhanes_histograms.png", hist_combined, width = 12, height = 8, dpi = 150)
cat("✔ Saved: nhanes_histograms.png\n")
```
# 5. Let us Determine the mean 60-second pulse rate for all participants in the data, the range of values for diastolic blood pressure in all participants and the variance and standard deviation for income among all participants?

```{r}
cat("\n=== DESCRIPTIVE STATISTICS ===\n")
# 4a. Mean 60-second pulse rate
mean_pulse <- mean(nhanes$Pulse, na.rm = TRUE) # No need to put na.rm = True, since we correct already misisng value sin our data
cat(sprintf("Mean 60-sec pulse rate : %.5f\n", mean_pulse))

# 4b. Range of diastolic blood pressure
bp_min <- min(nhanes$BPDia, na.rm = TRUE) # No need to put na.rm = True, since we correct already misisng value sin our data
bp_max <- max(nhanes$BPDia, na.rm = TRUE) # No need to put na.rm = True, since we correct already misisng value sin our data
cat(sprintf("Diastolic BP range     : %g – %g\n", bp_min, bp_max))

# 4c. Variance and SD for Income
income_var <- var(nhanes$Income,  na.rm = TRUE) # No need to put na.rm = True, since we correct already misisng value sin our data
income_sd  <- sd(nhanes$Income,   na.rm = TRUE) # No need to put na.rm = True, since we correct already misisng value sin our data
cat(sprintf("Income variance        : %.4f\n", income_var))
cat(sprintf("Income std deviation   : %.4f\n", income_sd))
```

# 6. Visualization the relationship between weight and height and coloration of the points by gender, diabetes and smoking status


```{r}
# Helper to make consistent scatter plots
scatter_wh <- function(data, color_var, title_suffix, palette = NULL) {
  p <- ggplot(data, aes(x = Height, y = Weight,
                        color = .data[[color_var]])) +
    geom_point(alpha = 0.45, size = 1.4) +
    labs(
      title    = paste("Weight vs Height —", title_suffix),
      x        = "Height (cm)",
      y        = "Weight (kg)",
      color    = color_var,
      caption  = "Source: CDC NHANES Dataset"
    ) +
    theme_nhanes +
    theme(legend.position = "bottom")
  
  if (!is.null(palette)) {
    p <- p + scale_color_manual(values = palette)
  } else {
    p <- p + scale_color_brewer(palette = "Set1", na.value = "grey70")
  }
  p
}

s1 <- scatter_wh(nhanes, "Gender",        "by Gender",
                 c("female" = "#E91E63", "male" = "#1565C0",
                   "Unknown" = "grey70"))

s1

s2 <- scatter_wh(nhanes, "Diabetes",      "by Diabetes Status",
                 c("Yes" = "#F44336", "No" = "#43A047",
                   "Unknown" = "grey70"))

s2

s3 <- scatter_wh(nhanes, "SmokingStatus", "by Smoking Status")

s3

scatter_combined <- (s1 | s2) / (s3 + plot_spacer()) +
  plot_annotation(
    title   = "NHANES — Weight vs Height",
    theme   = theme(plot.title = element_text(face = "bold", size = 16))
  )

print(scatter_combined)
ggsave("nhanes_scatter.png", scatter_combined, width = 14, height = 10, dpi = 150)
cat("✔ Saved: nhanes_scatter.png\n")
```


#7. T-test between the following variables and  conclusions on the relationship between them based on P-Value


## 7.1 Age and Gender

```{r}
# ── 6a. Age ~ Gender ─────────────────────────────────────────
age_female <- nhanes$Age[nhanes$Gender == "female"]
age_male   <- nhanes$Age[nhanes$Gender == "male"]

ttest_age_gender <- t.test(age_female, age_male)
cat("\n--- Age vs Gender ---\n")
print(ttest_age_gender)
cat(sprintf("Conclusion: p = %.4f → %s\n",
            ttest_age_gender$p.value,
            ifelse(ttest_age_gender$p.value < 0.05,
                   "SIGNIFICANT difference in age between genders.",
                   "NO significant difference in age between genders.")))
```

## 7.2 BMI and Diabetes

```{r}
# ── 6b. BMI ~ Diabetes ───────────────────────────────────────
bmi_yes <- nhanes$BMI[nhanes$Diabetes == "Yes"]
bmi_no  <- nhanes$BMI[nhanes$Diabetes == "No"]

ttest_bmi_diabetes <- t.test(bmi_yes, bmi_no)
cat("\n--- BMI vs Diabetes ---\n")
print(ttest_bmi_diabetes)
cat(sprintf("Conclusion: p = %.4e → %s\n",
            ttest_bmi_diabetes$p.value,
            ifelse(ttest_bmi_diabetes$p.value < 0.05,
                   "SIGNIFICANT difference in BMI between diabetic and non-diabetic participants.",
                   "NO significant difference in BMI between diabetic and non-diabetic participants.")))

```


## 7.3 Alcohol Year and Relationship Status

```{r}
# ── 6c. AlcoholYear ~ RelationshipStatus ─────────────────────
# Use the two largest groups for a clean two-sample t-test
rel_counts  <- sort(table(nhanes$RelationshipStatus), decreasing = TRUE)
cat("\nRelationshipStatus counts:\n"); print(rel_counts)

top2 <- names(rel_counts)[1:3]
alc_g1 <- nhanes$AlcoholYear[nhanes$RelationshipStatus == top2[1]]
alc_g2 <- nhanes$AlcoholYear[nhanes$RelationshipStatus == top2[3]]

ttest_alc_rel <- t.test(alc_g1, alc_g2)
cat(sprintf("\n--- AlcoholYear: '%s' vs '%s' ---\n", top2[1], top2[3]))
print(ttest_alc_rel)
cat(sprintf("Conclusion: p = %.4e → %s\n",
            ttest_alc_rel$p.value,
            ifelse(ttest_alc_rel$p.value < 0.05,
                   paste0("SIGNIFICANT difference in yearly alcohol consumption between '",
                          top2[1], "' and '", top2[3], "'."),
                   paste0("NO significant difference in yearly alcohol consumption between '",
                          top2[1], "' and '", top2[3], "'."))))
```




