---
title: "NHANES Public Health Data Analysis"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Show R code in the output
  message = FALSE,     # Suppress package messages
  warning = FALSE,     # Suppress warnings
  comment = "#>"       # Prefix for output lines
)
```

NHANES is a program run by the CDC to assess the health and nutritional
status of adults and children in the US. It combines survey questions
and physical examinations, including medical and physiological
measurements and laboratory tests, and examines a representative sample
of about 5,000 people each year. The data is used to determine the
prevalence of diseases and risk factors, establish national standards,
and support epidemiology studies and health sciences research. This
information helps to develop public health policy, design health
programs and services, and expand the nation's health knowledge.

You can find those data using these links:

**Link for the Data**: <https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv>

**Dictionary Data**: <https://github.com/HackBio-Internship/public_datasets/blob/main/R/nhanes_dd.csv>

**Tasks**:

• Process all NA (either by deleting or by converting to zero)

• Visualize the distribution of BMI, Weight, Weight in pounds (weight
\*2.2) and Age with an histogram.

• Determine the mean 60-second pulse rate for all participants in the
data?

• What’s the range of values for diastolic blood pressure in all
participants? (Hint: see help for min(), max()).

• What’s the variance and standard deviation for income among all
participants?

• Visualize the relationship between weight and height. Color the points
by:

o gender

o diabetes

o smoking status

• Conduct t-test between the following variables and make conclusions on
the relationship between them based on P-Value

o Age and Gender

o BMI and Diabetes

o Alcohol Year and Relationship Status

# 1. Load necessary packages

```{r}
# Install if not already installed
if (!require("pacman")) install.packages("pacman")

# Load all packages
pacman::p_load(
  tidyverse,   # data manipulation + ggplot2
  patchwork,   # combining plots
  scales       # axis formatting
)
```

```{r Packages}
library(tidyverse)
library(patchwork)
library(scales)
```

# 2. Load the Data and Present the Data Dictionnary

```{r}
# Source: CDC NHANES via HackBio public datasets repository
# Each row = one survey participant (~5,000 sampled per year)

# ── Load the data ──────────────
nhanes_raw <- read.csv(
  "https://raw.githubusercontent.com/HackBio-Internship/public_datasets/main/R/nhanes.csv",
  stringsAsFactors = FALSE   # keep character columns as strings, not factors
)

# ── Create the data dictionary as a data frame ──────────────
nhanes_dict <- data.frame(
  Column = c(
    "id", "Gender", "Age", "Race", "Education",
    "MaritalStatus", "RelationshipStatus", "Insured", "Income",
    "Poverty", "HomeRooms", "HomeOwn", "Work", "Weight", "Height",
    "BMI", "Pulse", "BPSys", "BPDia", "Testosterone",
    "HDLChol", "TotChol", "Diabetes", "DiabetesAge",
    "nPregnancies", "nBabies", "SleepHrsNight", "PhysActive",
    "PhysActiveDays", "AlcoholDay", "AlcoholYear", "SmokingStatus"
  ),
  Type = c(
    "integer",   "character", "integer",   "character", "character",
    "character", "character", "character", "numeric",
    "numeric",   "integer",   "character", "character", "numeric",  "numeric",
    "numeric",   "integer",   "integer",   "integer",   "numeric",
    "numeric",   "numeric",   "character", "integer",
    "integer",   "integer",   "integer",   "character",
    "integer",   "numeric",   "integer",   "character"
  ),
  Label = c(
    "Unique participant ID",
    "Biological sex of participant (male / female)",
    "Age in years at screening",
    "Race / ethnicity of participant",
    "Highest education level attained",
    "Current marital status",
    "Relationship status (Committed / Single)",
    "Whether participant has health insurance (Yes / No)",
    "Annual gross household income in USD",
    "Ratio of household income to poverty threshold",
    "Number of rooms in the participant's home",
    "Home ownership status (Own / Rent / Other)",
    "Current employment status",
    "Body weight in kilograms",
    "Standing height in centimetres",
    "Body Mass Index — weight(kg) / height(m)²",
    "60-second resting pulse rate (beats per minute)",
    "Systolic blood pressure (mmHg)",
    "Diastolic blood pressure (mmHg)",
    "Serum testosterone level (ng/dL)",
    "HDL (good) cholesterol level (mmol/L)",
    "Total cholesterol level (mmol/L)",
    "Whether participant was diagnosed with diabetes (Yes / No)",
    "Age at diabetes diagnosis (years)",
    "Total number of pregnancies",
    "Number of live births",
    "Average hours of sleep per night",
    "Whether participant is physically active (Yes / No)",
    "Number of days per week with physical activity",
    "Average number of alcoholic drinks on drinking days",
    "Number of days per year alcohol was consumed",
    "Smoking history (Never / Former / Current)"
  ),
  stringsAsFactors = FALSE
)

# ── View the full dictionary ─────────────────────────────────
print(nhanes_dict)

# ── Or view it as a clean table in RStudio Viewer ───────────
View(nhanes_dict)

# ── Quick lookup function — get label for any column name ───
get_label <- function(col_name) {
  match <- nhanes_dict$Label[nhanes_dict$Column == col_name]
  if (length(match) == 0) return("Column not found")
  return(match)
}

# Example usage:
get_label("BMI")          # "Body Mass Index — weight(kg) / height(m)²"
get_label("AlcoholYear")  # "Number of days per year alcohol was consumed"
get_label("BPDia")        # "Diastolic blood pressure (mmHg)"
get_label("Work")
```
# 3. Quick Overview of the data

```{r}
# Keep a working copy , never modify the raw object directly
nhanes <- nhanes_raw

# ── Structure of the dataset ─────────────────────────────────
# Shows column names, data types, and first values
str(nhanes)

# ── Name of all Variables ─────────────────────────────────
names(nhanes)

# ── Unique values in key categorical columns ─────────────────
# Useful to spot typos, unexpected categories, or hidden NAs
unique(nhanes$Gender)
unique(nhanes$Race)
unique(nhanes$Education)
unique(nhanes$MaritalStatus)
unique(nhanes$RelationshipStatus)
unique(nhanes$Diabetes)
unique(nhanes$SmokingStatus)
unique(nhanes$Work)
unique(nhanes$HomeOwn)

# ── First and last rows ──────────────────────────────────────
head(nhanes)          # first 6 rows
tail(nhanes, n = 10)  # last 10 rows

# ── Dimensions ───────────────────────────────────────────────
# How many participants (rows) and variables (columns)?
dim(nhanes)

# ── Detailed column-by-column overview ───────────────────────
# Shows data type and first values for every column
glimpse(nhanes)

# ── Statistical summary ──────────────────────────────────────
# Numeric columns → min, Q1, median, mean, Q3, max, NA count
# Character columns → length and class
summary(nhanes)

# ── Missing values per column ────────────────────────────────
# Shows exact NA count and percentage for each column

colSums(is.na(nhanes_raw))

na_count <- colSums(is.na(nhanes))
na_pct   <- round(na_count / nrow(nhanes) * 100, 2)

na_summary <- data.frame(
  Column       = names(na_count),
  Missing      = na_count,
  Percentage   = paste0(na_pct, "%")
)

na_summary

# Show only columns that actually have missing values
na_summary[na_summary$Missing > 0, ]
```



